---
title: "Iterative Refinement of Cellular Identity Using Online iNMF"
author: "Chao Gao and Joshua Welch"
date: "March 26, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Scenario 1: sampling minibatches from fully observed datasets
To perform online iNMF, we need to install the `online` branch. Please see the instruction below.
```{r install, message=FALSE, results='hide'}
library(devtools)
install_github("MacoskoLab/liger", ref = "online")
```
We first create a liger object by passing the filenames of HDF5 files containing the raw count data. The data can be downloaded [here](https://umich.box.com/s/m5cojytshxvk44pc4kki6clh6c9sv7ai).
```{r create.object, message=FALSE, results='hide'}
library(liger)
pbmcs = createLiger(list(stim = "stim_PBMCs.h5", ctrl = "ctrl_PBMCs.h5"))
```

We then perform the normalization, gene selection, and gene scaling in an online fashion, reading the data from disk in small batches.
```{r preprocess, message=FALSE, results='hide'}
pbmcs = normalize(pbmcs)
pbmcs = selectGenes(pbmcs, var.thresh = 0.2, do.plot = F)
pbmcs = scaleNotCenter(pbmcs)
```
## Online Integrative Nonnegative Matrix Factorization
Now we can use online iNMF to factorize the data, again using only minibatches that we read from the HDF5 files on demand (default mini-batch size = 5000).
```{r factorize, message=FALSE, results='hide'}
pbmcs = online_iNMF(pbmcs, k = 20, max.epochs = 5)
```
## Quantile Normalization and Downstream Analysis
After performing the factorization, we can perform quantile normalization to align the datasets.
```{r norm}
pbmcs = quantile_norm(pbmcs)
```

We can also visualize the cell factor loadings in two dimensions using t-SNE or UMAP.
```{r visualize}
pbmcs = runUMAP(pbmcs)
plotByDatasetAndCluster(pbmcs, axis.labels = c("UMAP1","UMAP2"))
```

We can also compare clusters or datasets (within each cluster) to identify differentially expressed genes. The runWilcoxon function
performs differential expression analysis by sampling a specified number of cells from the dataset on disk, then performing an in-memory Wilcoxon rank-sum test on this subset. Thus, users can still analyze large datasets with a fixed amount of memory.
```{r wilcox 1, results='hide'}
de_genes = runWilcoxon(pbmcs, compare.method = "clusters", max.sample = 5000)
```
```{r wilcox 2}
de_genes = de_genes[order(de_genes$padj), ]
head(de_genes[de_genes$group == 1,], n = 10)
```
##Scenario 2: iterative refinement by incorporating new datasets
We can also perform online iNMF with continually arriving datasets.
```{r prep1, message=FALSE, results='hide'}
MOp = createLiger(list(cells = "allen_smarter_cells.h5"))
MOp = normalize(MOp)
MOp = selectGenes(MOp, var.thresh = 2)
MOp.vargenes = MOp@var.genes
MOp = scaleNotCenter(MOp)
```

```{r fact1, message=FALSE, results='hide'}
MOp = online_iNMF(MOp, k = 40, max.epochs = 1)
```

```{r vis1}
MOp = quantile_norm(MOp)
MOp = runUMAP(MOp)
plotByDatasetAndCluster(MOp, axis.labels = c("UMAP1","UMAP2"))
```

```{r prep2, message=FALSE, results='hide'}
MOp2 = createLiger(list(nuclei = "allen_smarter_nuclei.h5"))
MOp2 = normalize(MOp2)
MOp2@var.genes = MOp@var.genes
MOp2 = scaleNotCenter(MOp2)
```

```{r fact2, message=FALSE, results='hide'}
MOp = online_iNMF(MOp, X_new = list(nuclei = "allen_smarter_nuclei.h5"), k = 40, max.epochs = 1)
```
```{r vis2}
MOp = quantile_norm(MOp)
MOp = runUMAP(MOp)
plotByDatasetAndCluster(MOp, axis.labels = c("UMAP1","UMAP2"))
```

##Scenario 3: projecting new datasets
```{r prep3, message=FALSE, results='hide'}
MOp = createLiger(list(cells = "allen_smarter_cells.h5"))
MOp@var.genes = MOp.vargenes
MOp = online_iNMF(MOp, k = 40, max.epochs = 1)
MOp = quantile_norm(MOp)
MOp = runUMAP(MOp)
plotByDatasetAndCluster(MOp, axis.labels = c("UMAP1","UMAP2"))
```

```{r project, message=FALSE, results='hide'}
MOp = online_iNMF(MOp, X_new = list(nuclei = "allen_smarter_nuclei.h5"), k = 40, project = TRUE)
```

```{r vis3}
MOp = quantile_norm(MOp)
MOp = runUMAP(MOp)
plotByDatasetAndCluster(MOp, axis.labels = c("UMAP1","UMAP2"))
```








